---
title: "final_project"
date: last-modified
execute: 
  eval: true
  warning: false
  message: false
format:
  html:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(ggplot2)
library(here)
library(kableExtra)
library(car)
library(broom)
```

```{r}
food <- read_csv(here("data","food_affordability_2006_2010.csv"))
counties <- read.csv(here("data","california_counties.csv"))
```

#### Data Cleaning 

```{r}
food <- food %>%
 filter(!is.na(median_income)) %>% # removed all median_income NAs as they are crucial for lm 
  select(-region_name) # removed MPO region names
```

```{r}
# Create a histogram for median_income
income_hist <- ggplot(food, aes(x = median_income)) +
  geom_histogram(binwidth = 5000, fill = "firebrick2", color = "white") +
  labs(title = "Histogram of median income", 
       x = "Median income", 
       y = "Frequency") +
  theme_bw() +
  theme(text = element_text(family = "courier"),
        panel.background = element_rect(colour = "skyblue")) 

income_hist
```

```{r}
summary(food$cost_yr) %>%
  kable() %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = "bordered",
                html_font = "courier") %>%

summary(food$median_income) %>%
  kable() %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = "bordered",
                html_font = "courier") %>%

food <- food %>% 
  filter(median_income < 125000) 

```

```{r}
# classify counties by region to make 5 level categorical variable
# if I had used the region_name col from original data the categorical variable would have 15 levels, as opposed to 5 if I follow the MPO region designations (Metropolitan Planning Organization)

superior <- c("Butte","Colusa","Del Norte","Glenn","Humboldt","Lake","Lassen","Mendocino","Modoc","Nevada","Plumas","Shasta", "Sierra","Siskiyou","Tehama","Trinity")

central <- c("Alpine","Amador","El Dorado","Fresno","Inyo","Kings","Madera","Mariposa","Merced","Mono","Placer","Sacramento","San Joaquin","Stanislaus","Sutter","Yuba","Tulare","Tuolumne","Yolo", "Calaveras")

bay_area <- c("Almeda","Contra Costa","Marin","Monterey","Napa","San Benito","San Fracisco","San Mateo","Santa Clara","Santa Cruz","Solano","Sonoma")

southern <- c("Imperial","Kern","Orange","Riverside","San Bernardino","San Diego","San Luis Obispo","Santa Barbara","Ventura")

los_angeles <- "Los Angeles"

region_mapping <- data.frame(
  county_name = c(superior, central, bay_area, southern, los_angeles),
  region_name = rep(c('superior', 'central', 'bay_area', 'southern','los_angeles'), 
               times = c(length(superior), length(central), length(bay_area), length(southern),length(los_angeles))))

food <- inner_join(food,
                   region_mapping,
                   by = "county_name")
```

```{r results='asis'}
cost_fam_lm <- lm(cost_yr ~ median_income + region_name + ave_fam_size + region_name * median_income, data = food)
# should try removing the outliers in median_income and affordability food ratio
# in "ave_fam_size" all races are combined, race variable not ideal in model

cost_fam_summary <- tidy(cost_fam_lm) %>%
  kable(col.names = c("Variable","Estimate","Std.Error", "Statistic","p.value")) %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = "bordered",
                html_font = "courier") %>%
  print()
```

#### Run a VIF to detect multicollinearity, if present
```{r results='asis'}
vif_results <- vif(cost_fam_lm) 
vif_results %>%
  as.data.frame()

vif_table <- vif_results %>%
  kable("html", col.names = c("Variable", "GVIF", "Degrees of freedom", "Adjusted GVIF")) %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = "bordered",
                html_font = "courier") %>%
  print()
```

High level of multicollinearity between region and food cost, and between region and median_income but when adjusted for degrees of freedom, not severe enough to drop the predictors with the high VIF


# response variable = affordability ratio
# independent variables / predictors
# county - but must drop NAs (136) <- categorical
# median_income <- numerical 
# average family size <- numerical 

#### Hypothesis:
median income, family size 
female-headed households



